---
title: "AD PROJECT 2: Prediction of male body fat from body measurements"
output: html_notebook
---
*Lucía De Pineda & Sonia Castro* , *Group 12*

The first thing we do is load the needed libraries:
```{r}
library(GGally)
library(openintro)
library(tidyverse)
library(factoextra)
library(ggplot2)
library(dplyr)
library(factoextra)
library(readr)
```

Now we save our dataset into a dataframe and we print the first columns to have an idea of the values for each variable:
```{r}
data<-read.table(text = gsub("  ", " ", readLines("data.txt")), header=TRUE, dec=".")
head(data)
```

We check the dimension of the data:
```{r}
dim(data)
```

We check the correlation of the variable density with the target variable:
```{r}
ggpairs(data[,c(1,2)])
```

We decide to drop the variable density:
```{r}
data <- data[2:15]
head(data)
```


# PCA

```{r}
z = cor(data)
```

```{r}
z[lower.tri(z,diag=TRUE)]=NA  # Prepare to drop duplicates and meaningless information
z=as.data.frame(as.table(z))  # Turn into a 3-column table
z=na.omit(z)  # Get rid of the junk we flagged above
z=z[order(-abs(z$Freq)),]    # Sort by highest correlation (whether +ve or -ve)
z
```
The highest correlation is between Weight and Hip (0.9409). The one with highest correlations is Weight.

Highest correlations with target variable Fat are Abdomen and Chest.
```{r}
ggpairs(data[,c(1,6:7)])
```

Then we compute the PCA for our data, without using our target variable.
```{r}
data2 = data[,2:14]
data2 = scale(data2)
data_pca = prcomp(data2)
```

We check the proportions of variability explain by the PCs:
```{r}
(sdev_pca=data_pca$sdev)
(var_pca=sdev_pca^2)
```
```{r}
var_total=sum(var_pca)
var_total
```
```{r}
(prop1=var_pca[1]/var_total)
(prop2=var_pca[2]/var_total)
(prop3=var_pca[3]/var_total)
(prop4=var_pca[4]/var_total)
prop1+prop2+prop3+prop4
```

Representation of the percentages:
```{r}
fviz_eig(data_pca)
```

We print the PCs:
```{r}
data_pca$rotation
```

Plot of the first four PCs:
```{r}
fviz_pca_var(data_pca, axes = c(1,2))
fviz_pca_var(data_pca, axes = c(3,4))
```

Dividing category Fat into groups to see it in the biplot:
```{r}
hist(data$X.Fat)
```

```{r}
(Fat.cat = cut(data$X.Fat, breaks = seq(0, 50, 10)))
```


```{r}
levels(Fat.cat) = c('0','1','2','3','4')
```

```{r}
data_cat <- data
data_cat$Fat <- Fat.cat
```

We do the biplot:
```{r}
options(ggrepel.max.overlaps = 10)
fviz_pca_biplot(data_pca, repel=TRUE, axes = c(1,2), habillage = data_cat$Fat)
```

We check the values that stand out:
```{r}
(val_39=data2[39,])
t(data_pca$rotation)%*%val_39
```


```{r}
(val_41=data2[41,])
t(data_pca$rotation)%*%val_41
```
 

```{r}
(val_42=data2[42,])
t(data_pca$rotation)%*%val_42
```


# MDS

First we compute the Euclidean distance for each pair of rows, creating the distance table $D$.
```{r}
D=dist(data2, method = "euclidean")
```

Then we perform MDS and we check the proportions of variance explained:
```{r}
mds=cmdscale(D,k=4,eig=TRUE)
pve = mds$eig/sum(mds$eig)
head(pve)
```

Plot of two first axes of PCA:
```{r}
mds<-as.data.frame(mds$points)
ggplot(mds,aes(x=V1,
                y=V2,
                label=rownames(data)))+
  geom_text(alpha=0.8,size=3,col="salmon")
```

We calculate the distance matrix of the MDS and then the residual matrix:
```{r}
D2=dist(mds,method="euclidean")
res=D-D2
as.matrix(res)
```

For reference we take the mean value of the residual matrix:
```{r}
mean(res)
```

Histogram of the residuals:
```{r}
hist(res, breaks=seq(0,5.5,0.25))
```


# CLUSTER

### K-means

First, we choose the value of k:
```{r}
# TWSS to choose k
fviz_nbclust(data2, kmeans, method = "wss")
```

We choose k=4, then we run kmeans and we plot the cluster
```{r}
# run k-means
km = kmeans(data2, centers= 4)
# cluster assignment + PCA scores
fviz_cluster(km, data2, geom = "point")

```


```{r}
PCs = prcomp(data2)
fviz_pca_biplot(PCs, geom = "text", habillage = as.numeric(km$cluster))
```



## Agglomerative hierarchical


Choosing k:
```{r}
fviz_nbclust(data2, FUN=hcut, hc_method = "complete", method = "wss")
```


We choose k=5. Then we choose the linkage method and we plot the cluster assignment:
```{r}
D = dist(data2)
hc = hclust(D, method = "complete")
cl = cutree(hc, k = 5)
fviz_cluster(list(data = data2, cluster = cl), geom = "point")
```

We plot the dendogram:
```{r}
fviz_dend(hc, k=5)
```

Finally we plot the cluster assignment with the PCs scores:
```{r}
PCs = prcomp(data2)
fviz_pca_biplot(PCs, geom = "text", habillage = as.numeric(cl))
```


